---
title: "Problem Set 2 - Data and Programming for R"
author: "Meredith Gavin"
date: "29/03/2020"
output:
  pdf_document: default
  html_document:
    number_sections: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

<!-- .Rmd files use  markdown, a text mark up language, to provide formating.--> 
<!--Text include within these strange arrows are comments and will not show up when you knit-->

# Front matter
This submission is my work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: **MG**

Add your collaborators: **__**

Late coins used this pset: 0. Late coins left: 5 
<!--You may use up to two for a given assignment.)-->

# Problems
## You change section heading goes here 
<!--(Notice the use of two `##`)-->

1. YOUR TEXT ANSWER GOES HERE (Please delete)


# 2 GIT CONCEPTS

## 1
4 benefits of distributed version control:

-collaborators can work together on an R project without overlapping or accidentally deleting each other's code
-it tracks changes to code, so that you can better follow edits made to your code
-Organization is a concurrent part of the system, as opposed to its own separate task
-committing your code makes it less likely to be lost or corrupted

Resource: https://peerj.com/preprints/3159v2/

## 2
### i

The remote repository for this project is datasci-harris/skills-problem-set-2-meredithgavin

### ii

To add a file to staging in github, clone it from the remote repository and then open in an editor. 

### iii

To commit an issue to the local repository, run git commit or (if its linked with something like R Studio) save the file and changes will show up in github desktop. 

### iv 

github desktop show the most recently saved changes to the code which have not yet been committed to the remote repository.

### v

I am on the "main" branch because I am wroking individually on this problem set. 

### vi

#### a

Once you create a new branch, you can begin working on that branch in the working directory.

#### b

The new branch will not be available in the remote repository until it is committed and pushed to the remote repository from github desktop. 

#### c

We may want to start a new branch to separate our work from a partner's work before its committed to the main branch.


# 3.1: PRACTICING WITH DPLYR VERBS

## 1

```{r}
my_variable <- 10
my_variable
```


## 2
```{r}
library(tidyverse)
ggplot(mpg, mapping = aes(x=displ, y=hwy)) + geom_point()
filter(mpg, cyl == 8)
filter(diamonds, carat > 3)
```


## 3

The Keyboard Shortcut Quick Reference pops up. You can also find the guide by going to Tools -> Keyboard Shortcut Help. 

# 3.2  TRANSFORMATIONS: FILTER ROWS WITH FILTER ()

## 1

```{r}
library(nycflights13)
flights <- nycflights13::flights
head(print(flights))
```

### i

```{r}

head(filter(flights, arr_delay >= 3))
```
### ii

```{r}
head(filter(flights, dest == "IAH" | dest == "HOU"))
```
### iii
```{r}
head(filter(flights, carrier == "UA" | carrier == "AA" | carrier == "SW"))
```
### iv

```{r}
head(filter(flights, month ==3  | month == 4 | month == 6 ))
```

### v

```{r}
head(filter(flights, dep_delay <= 0 & arr_delay >= 2))
```

### vi

```{r}
speedbonus <- flights$arr_delay - flights$dep_delay

head(filter(flights, dep_delay >= 60 & speedbonus > 30))
```

### vii

```{r}
head(filter(flights, dep_time <= 500))
```



## 2

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes (x = displ, y = cty, color = drv, shape = drv)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = cty), se=FALSE)

filter(mpg, displ < 4 & displ > 3 & cty < 13)
```

The green triangle is a 2008 dodge caravan with two wheel drive. 


### 1

```{r}

filter(flights, is.na(arr_time))


```

### 2

```{r}
flights %>%
  filter(is.na(dep_time)) %>%
  summarise(n=n())

head(flights %>%
       filter(is.na(dep_time)))
```

There are 8255 flights with missing departure times. Of these flights, there are
also missing values for air_time, arr_delay, arr_time, and dep_delay. These rows represent information that can only be gathered for flights that departed. 

### 3 

NA|TRUE returns observations of values that are missing OR values that are true given the parameters. 

### 4 

FALSE & NA returns observations with values that are False AND NA. 

# 3.3: TRANSFORMATIONS: SELECT COLUMNS WITH SELECT()

## 1
```{r}
head(select(flights, month, day, year))
```

Selecting multiple variables will return the dataframe with only the selected variables' columns. 

## 2    

```{r}
select(flights, contains("TIME", ignore.case = FALSE))

# Source: https://stackoverflow.com/questions/54634124/dplyrend-with-and-case-sensitivity
```



## 3

```{r}
head(select(flights, dep_time, dep_delay, arr_time, arr_delay))

new_df <- data.frame(flights$dep_time, flights$dep_delay, flights$arr_time, flights$arr_delay)
head(new_df)
  
```


# 3.4: TRANSFORMATIONS: SORT COLUMNS WITH ARRANGE()

## 1

```{r}
arrange(flights, desc(arr_delay)) %>% 
  head(1)
```

The most delayed flight by arrival time (the largest value of arr_delay) was flight N384HA from JFK to HNL. It arrive 1272 minutes late. 


## 2

```{r}
arrange(flights, dep_delay) %>% 
  select(tailnum, day, month, dep_delay) %>%
         head(5)
```


## 3

```{r}
arrange(flights,! is.na(dep_delay), dep_delay)

# Resource: https://stackoverflow.com/questions/25258817/how-to-have-nas-displayed-first-using-arrange
```

# 3.5: TRANSFORMATIONS: ADD NEW VARIABLES WITH MUTATE()

## 1

```{r}
library(tidyr)
flight_min <- flights %>% 
 mutate(arr_time_min = (arr_time/100) *60) %>%
  mutate(dep_time_min = (dep_time/100) *60)

head(flight_min)
```

## 2

```{r}
ggplot(flight_min, mapping = aes(x=air_time, y= (arr_time_min - dep_time_min))) +
  geom_point()
```

## 3

### i
```{r}
flight_min %>%
  mutate(difference = arr_time_min - dep_time_min) %>%
  filter(!("air_time" == "difference")) %>%
  head(5)
```
The data is indicating that none of the flights have the same air time as they have difference in arrival and departure time.

### ii

One reason for this may be the number of decimal points that R defaults to. The difference column defaults to a high level of specifity. As a result, R reads the two as always unequal.

### iii

NA values may interfere with the filter code. 

### iv

One of the values may be an integer while the other may be a character. 


## 4

```{r}
flight_min <- flight_min %>%
  mutate(arr_time_min = (as.integer(arr_time_min))) %>%
  mutate(dep_time_min = (as.integer(dep_time_min))) %>% 
  mutate(air_time = as.integer(air_time)) %>%
  mutate(difference = (as.integer(arr_time_min-dep_time_min)))

flight_min %>%
  filter(!(air_time == difference))

head(flight_min)

ggplot(flight_min, mapping = aes(x=air_time, y= difference)) +
  geom_point()
```


## 5

Applying as.integer eliminated the extra decimals and normalized the type of vector across the compared variables. However, the code is returning no flights for which the air_time value and the difference value are not equal. 

## 6

```{r}
arrange(flights, dep_delay) %>%
  head(10)

min_rank(flights$dep_delay) %>%
  head(5)
```

One reason there may by more than 10 flights in the top ten is that some flights have equal delays. 

# 3.6: TRANSFORMATIONS: SUMMARIZING GROUPED DATA


## 1

```{r}
not_cancelled <- flights %>%
  filter(! is.na(arr_time))

head(not_cancelled)

not_cancelled %>%
  summarise(dest=n()) %>%
  summarise()

not_cancelled %>% count(tailnum, wt
= distance) 

not_cancelled %>%
  group_by(tailnum) %>%
  summarise(Freq = sum(distance))

#Resource: https://www.statology.org/sum-by-group-in-r/
```


There are 328063	observations of dest in not_cancelled. 

## 2


```{r}
flights_by_dest <- flights %>%
  filter(origin=="LGA"|origin=="JFK") %>%
  group_by(dest)

mutate(flights_by_dest,
       mean_arr_delay = (mean(arr_delay, na.rm=TRUE))) %>%
  arrange(desc(mean_arr_delay)) %>%
  head(5)
```

The destination with the highest average arrival delay from NYC is CAK.


## 3

```{r}
flights %>%
  group_by(hour) %>%
  arrange(arr_delay) %>%
  head(5)
```

The best hours of the day to leave to avoid arrival delays are 5pm and 7am.

## 4

```{r}
flights_by_tailnum <- flights %>%
  group_by(tailnum) %>%
  mutate(total_delay = arr_delay + dep_delay)

flights_by_tailnum %>%
  arrange(desc(total_delay)) %>%
  select(tailnum, total_delay)
```
N384HA has the longest total delays.


## 5

```{r}
flights %>%
  group_by(dest) %>%
  summarise(n=n_distinct(carrier, na.rm = TRUE)) %>%
  filter(n >= 3)

#Resource: https://stackoverflow.com/questions/56999897/sum-of-unique-values-in-a-column
```

## 6

```{r}
not_cancelled %>%
  group_by(carrier) %>%
  summarise(dep_time = n())
```

## 7
```{r}
flights %>%
  filter(origin=="LGA"|origin=="JFK") %>%
  group_by(carrier) %>%
  count(dest) %>%
  filter(n == 1)
  

```
5 carriers that originate in NYC only fly to one destination. 

# 3.7: TRANSFORMATIONS: PRACTIVAL APPLICATION 

## 1

```{r}
flights %>%
  group_by(carrier) %>%
  summarise(mean(arr_delay, na.rm=TRUE)) %>%
  head(5)
```


### a

```{r}
flights %>%
  group_by(carrier) %>%
  summarise(median(arr_delay, na.rm=TRUE)) %>%
  head(5)
```



### b

One way to identify airlines with higher average arrival delays is to create a visual representation of the average delay by airline. 

Another way is to find the total mean of arrival delays, filter for arrival delays above the mean and identify carriers who appear most often. 

### c

```{r}
flights_delay <- flights %>% 
  group_by(carrier) %>% 
  summarise(me = mean(arr_delay,na.rm = TRUE))

ggplot(flights_delay, mapping = aes(x = carrier, y = me)) + 
  geom_bar(stat = "identity")

#Resource: https://stackoverflow.com/questions/33096781/how-to-create-a-bar-plot-and-show-average-y-values

```

```{r}
flights %>%
  group_by(carrier) %>%
  filter(arr_delay > mean(arr_delay, na.rm = TRUE)) %>%
  head(5)
  
```


# 3.8: TRANSFORMATIONS: GROUP OPERATIONS WITH MUTATE() AND FILTER() 

## 1

### a
```{r}
flights_priordelays <- flights %>% 
  arrange(origin, dest, day, sched_dep_time) %>%
  group_by(origin) %>%
  mutate(prior_delay = lag(dep_delay))

head(flights_priordelays)

# Resource: https://dplyr.tidyverse.org/reference/lead-lag.html

```

### b

```{r}
flights_priordelays %>%
  group_by(prior_delay) %>%
  ggplot(mapping = aes(x=mean(dep_delay, na.rm= TRUE), y=prior_delay)) +
  geom_point()
```


## 2

### a

```{r}
flights %>%
  group_by(dest) %>%
  mutate(med = median(air_time, na.rm=TRUE)) %>%
  ungroup() %>%
  filter(air_time > med) %>%
  mutate(diff_air_time = air_time - med) %>%
  arrange(desc(diff_air_time)) %>%
  head(5)

```


## 3

```{r}
flights %>%
  group_by(tailnum) %>%
  mutate(total_delay = cumsum(dep_delay >= 60), na.rm=TRUE) %>%
  mutate(total_delayed_flights = sum(total_delay > 1)) %>%
  head(5)
```








