---
title: "Skills PS4"
author: "Meredith Gavin"
date: "4/30/2022"
output:
  pdf_document: default
  html_document:
    number_sections: yes
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)

```

<!-- .Rmd files use  markdown, a text mark up language, to provide formating.--> 
<!--Text include within these strange arrows are comments and will not show up when you knit-->

**Front matter**
This submission is my work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: **MG**

Add your collaborators: **__**

Late coins used this pset: 1. Late coins left: 4. 


# Problems
## You change section heading goes here 
<!--(Notice the use of two `##` followed by a space. )-->



# 1: Tidy Data with pivot_longer() and pivot_wider()


## 1.1 

Load the tables: 

```{r}
library(tidyverse)

?table1

table1 <- table1
head(table1)


table2 <- table2
head(table2)

table4a <- table4a
head(table4a)

table4b <- table4b
head(table4b)
```

Calculate the TB rates:

Using table1:

```{r}
table1 %>% 
  mutate(tb_rate = (cases/population)*10000)
```

Using table2:

```{r}
table2_with_rates <- table2 %>%
  pivot_wider(
    names_from = type,
    values_from = count
  ) %>%
  mutate(tb_rate = (cases/population)*10000)

head(table2_with_rates)
```

Using table4a and table4b:

First, we needed to adjust the tables so that they could be compatible for a
join. Once the tables are compatible, we can join table4a and table4b to
create a table4. Table4 includes all the information we need to compute
infection rates. 

```{r}
table4a_cleaned <- pivot_longer(table4a,
             cols = `1999`:`2000`,
             names_to = "year",
             values_to = "cases"
             )

table4b_cleaned <- pivot_longer(table4b,
             cols = `1999`:`2000`,
             names_to = "year",
             values_to = "population")

table4 <- table4a_cleaned %>%
  left_join(table4b_cleaned,
            by = c("year", "country"))

table4 %>%
  mutate(tb_rate = (cases/population)*10000)
```

Table 1 was the easiest to work with because the table didn't need to be 
adjusted at all in order to create the tuberculosis rate variable. Tables 4a
and 4b were the most difficult to work with because neither had the full 
information necessary. As a result, we had to adjust the tables to make them
compatible and then join the two before creating the tuberculosis rate 
variable. 


## 1.2

To recreate the plot from section 12.2 using table2 instead of table1, we
can use table2_with_rates, which is is table2 with TB rates. From there, we
can use the table to calculate the number of cases each year using count(). 
Then we can use the code for plotting table1 to plot table2_with_rates. 

```{r}
head(table2_with_rates)

table2_with_rates %>%
  count(cases,
        wt = year) %>%
  ggplot(data = table2_with_rates, 
         mapping = aes(year, cases)) + 
  geom_line(aes(group = country), 
            colour = "grey50") + 
  geom_point(aes(colour = country))
```



## 1.3

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
) 

stocks


 stocks %>%
  pivot_wider(names_from = year, 
              values_from = return) %>%
  pivot_longer(`2015`:`2016`,
               names_to = "year", 
               values_to = "return")
```

The "names_from =" and "values_from =" arguments do not need quotation marks
because they are referring to the integer values for year and return. The
"names_to =" and "values_to =" arguments need quotations because we want R
to recognize them as header names, not integer values. 


## 1.4


```{r}
library(reprex)

reprex({
  library(tidyverse)
  table4a %>% pivot_longer(1999:2000, 
                         names_to = "year", 
                         values_to = "cases")})
```


Reprex Output: 

library(tidyverse)
table4a %>% pivot_longer(1999:2000, 
                       names_to = "year", 
                       values_to = "cases")
 Error in `loc_validate()`:
  Can't subset columns past the end.
  Locations 1999 and 2000 don't exist.
  There are only 3 columns.



## 1.5

The reprex output indicates that R is having difficulty identifying the
column headers "1999" and "2000." As a result, the function cannot create a
data table as requested. The code is identifying the years 1999 and 2000 as
integers instead of as character values. We can fix this by writing the
years in single quotation marks.

```{r}
table4a %>% pivot_longer(`1999`:`2000`, 
                         names_to = "year", 
                         values_to = "cases")
```

## 1.6

```{r}
people <- tribble(
~name, ~key, ~value, 
#-----------------|--------|------

"Phillip Woods",    "age",      45,
"Phillip Woods",    "height",   186,
"Phillip Woods",    "age",      50,
"Phillip Woods",    "height",   185,
"Jessica Cordero",  "age",      37,
"Jessica Cordero",  "height",   156
)

people %>%
  mutate(unique_id = row_number()) %>%
  pivot_wider(names_from = key,
              values_from = value)
```

The original problem with the tribble is that we have four observations of
the name "Phillip Woods" and 2 observations of each age and height. When we 
pivot_wider(), R understands all 4 "Phillip Woods" as the same observation, 
however that doesn't work because one person cannot have two heights and two
ages. To fix this, we add a unique identifier column. This allows R to read 
each name in the tribble as an individual observation. We could have
multiple people with the same name, so we can't assume that any height or
age is paired. R then fills in unknown observations with NA. 


## 1.7

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes", NA, 10,
  "no", 20, 12
)

preg

preg %>%
  pivot_longer(male:female,
               names_to = "sex",
               values_to = "count") %>%
  pivot_wider(names_from = pregnant,
              values_from = count) %>%
  rename(pregnant = yes,
         not_pregnant = no)

#Resource: https://www.datasciencemadesimple.com/rename-the-column-name-in-r-using-dplyr/
```


I used both pivot_longer() and pivot_wider(). Using pivot_longer(), I
created a tibble where "pregnant" was still a column with two observations
of yes and two observations of no (to correspond with two observations each
of male and female). From there, I used pivot_wider() to create columns
based on the values yes and no in the "pregnant" column. For ease of
understanding, I renamed the "yes" column to pregnant and the "no" column to
not_pregnant. The variables are now sex, pregnant, and not pregnant. 


## 1.8

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"),
           extra = "merge")

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"),
           extra = "merge") 
```

There are three options for "extra =", warn, drop, and merge. If there are too 
many values in the tibble, extra tells R what to do with it when separating the
observations in columns. Warn, which is the default, tells R to warn us when 
there is an extra value. Drop gets rid of extra values without warning. Merge
adds the value into the table by creating a cell with more than one 
observation. 

Fill tells R what to do with a missing value when converting a tibble into a 
table. In this case there is one less value in column two than in columns one
and three. The default, fill = "warn", creates a warning when there is a 
missing value. Right fills the empty spot with the observation to the right, 
and left fills the empty value with the observation from the left. 


## 1.9 

```{r}
who <- tidyr::who

who %>%
  select(country, iso2, iso3) %>%
  group_by(country) %>%
  distinct() %>%
  filter(n() > 1)
```

By grouping isolating the country, iso2, and iso3 columns and then grouping
by country, we create what should be a list of unique rows. Using filter(), we
can check for any repeated rows. Because there are no repeated rows, we can 
infer that country, iso2, and iso3 are individual country names and codes. 

Also, the data is pulled from the World Health Organization (WHO). The 
organization provides a data dictionary dataset which confirms that iso2 and 
iso3 are both country identifiers. 



# 2 Tidying Case Study

Load the data from R4DS

```{r}

#The original dataset:

who <- who
head(who)

#The tidied dataset:
who_tidy <- who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65, 
    names_to = "key", 
    values_to = "cases", 
    values_drop_na = TRUE
  ) %>% 
  mutate(
    key = stringr::str_replace(key, "newrel", 
                               "new_rel")
  ) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)

head(who_tidy)
```


## 2.1 


### 2.1.a

The textbook sets values_drop_na = TRUE to get rid of NA values in the data.
If we dropped that part of the code, the data would look like this:

```{r}
who_keep_na <- who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65, 
    names_to = "key", 
    values_to = "cases"
  ) %>% 
  mutate(
    key = stringr::str_replace(key, "newrel", "new_rel")
  ) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)

head(who_keep_na)

head(who_tidy)
```


We can use complete() to identify missing values:

```{r}
who_tidy %>%
  nrow()
who_tidy %>%
  complete(country, year, sex, age) %>%
  nrow()

132653 - 76046
```

Using complete(), we completed the dataset with implicitly missing 
observations. For example, we simply didn't have an observation of 
Afghanistan's tuberculosis cases in 1987. Once we have that information, 
we can count the total number of rows and compare it to the number of rows
in who_tidy. In total, we have 5,6607 missing observations when compared
with who_tidy in which NA values have been removed. 

### 2.1.b 

```{r}
who_keep_na %>%
  filter(is.na(country) | is.na(year))

```


The tidied who data with the NA values kept is not explicitly missing any
country-year pairs. 

## 2.2

In the WHO case study, an NA refers to a value which they simply have no 
data on. For example, if there is no information on tuberculosis cases in 
Afghanistan in 1983, the cases would be coded as NA. Zero refers to an 
observation of zero. If they do have data on tuberculosis in Afghanistan in
1983, but there were no cases, then it would be coded as '0'. 

## 2.3 

```{r}

#The tidied dataset removing the mutate() step:
who_tidy_no_mutate <- who %>%
  pivot_longer(
    cols = new_sp_m014:newrel_f65, 
    names_to = "key", 
    values_to = "cases", 
    values_drop_na = TRUE
  ) %>% 
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)

head(who_tidy_no_mutate)
head(who_tidy)
```

The mutate() code tells R to replace the name "newrel" with "new_rel". When
we don't do that, we get an error in which R identifies missing values in 
2580 rows and replaces them with "NA". In reality, the values should 
(mostly) be available, its just an issue of technical differences in the
names. 


## 2.4

### 2.4.a

The total number of cases (n_cases) by sex for each country-year pair:

```{r}
who_tidy_grouped <- 
  who_tidy %>%
  group_by(year, country, sex) %>%
  summarise(n_cases = sum(cases, na.rm = TRUE))

head(who_tidy_grouped)
```

### 2.4.b

Raw values likely won't provide clear evidence because we would need to 
individually compare case counts between men and women in each country over
each year of observation. Some countries only have a small number of cases 
each year, which is not enough to create solid evidence of significant 
differences in the number of cases between men and women. We need a separate
variable that allows us to compare differences across countries and years. 

### 2.4.c

The ratio of male to female patients with tuberculosis (m_to_f_ratio) for
each country-year pair:

```{r}
who_tidy_ratios <-
  who_tidy %>%
  group_by(year, country, sex) %>%
  summarise(n_cases = sum(cases, na.rm = TRUE)) %>%
  pivot_wider(names_from = "sex", values_from = "n_cases") %>%
  mutate(m_to_f_ratio = m/f, 
         n_cases = f + m)

head(who_tidy_ratios)
```


### 2.4.d

Grouping the table only by year would likely skew the data. Some countries
have very few cases while others have a very high number, certain countries
may be hot spots in certain years at the same time that others are seeing a
lull in case counts, and country cultures may impact the ratio. For example,
in countries where women are more likely to be caring for the sick, they may
see higher rates of women being diagnosed with tuberculosis. 

### 2.4.e

```{r}

age_labels <- c(
  '014' = "0-14",
  '1524' = "15-24",
  '2534' = "25-34",
  '3544' = "35-44",
  '4554' = "45-54",
  '5564' = "55-64",
  '65' = "65+"
)

head(who_tidy)

who_tidy %>%
  group_by(country, year, sex, age) %>%
  summarise(n_cases = sum(cases, na.rm = TRUE)) %>%
  group_by(country, year) %>%
  mutate(total_cases = sum(n_cases)) %>%
  ungroup() %>%
  mutate(percent_by_sex = n_cases/total_cases) %>%
  filter(total_cases > 1,
         year >= 1997) %>%
  ggplot(mapping = aes(y = year,
                       x = percent_by_sex,
                       color = sex)) +
  geom_point(aes(alpha = 0.2)) +
 facet_grid(age ~., labeller = as_labeller(age_labels)) +
  #Resource: https://stackoverflow.com/questions/3472980/how-to-change-facet-labels
  labs(y = "Year",
       x = "Percentage of Cases",
       color = "sex",
       title = "Percentage of Tuberculosis Patients by Sex",
       subtitle = "1997 to 2013",
       tag = "Age") +
  theme(plot.tag.position = c(.88, 0.5),
        plot.tag = element_text(angle = -90))

 # Resource: https://stackoverflow.com/questions/59632276/how-to-add-vertical-label-on-the-right-side-of-ggplot-graph

```

### 2.4.f

The graph includes information on percentage of cases by sex each year,
grouped by age. I filtered the data to include only observations where 
cases are greater than 1, to eliminate scenarios in which percentage is
100% one sex only because there is only on tuberculosis case. The plot 
focuses only on cases 1997 onward, and it indicates a difference in the
percentages of male and female cases. There is no clear grouping of 
male and female patients in the 0 to 14 age bracket. This could be because 
there is a relatively low number of cases in this age group. In the 25-34 
and 35-44 age brackets, the percentage of male cases clearly outweigh the 
percentage of female patients. By the time we reach the 65+ age bracket,
the pattern is less defined, but it does appear that male patients still
make up a larger proportion of total cases. It is less defined, however, 
perhaps because there is a greater number of cases in elderly populations
and/or because women tend to outlive men, so women make up a larger portion 
of that age bracket to begin with. It was difficult to plot this data and to
present it in a way that is representative of the data. In this scenario, I
omitted country distinctions in the plot. There were simply too many
countries to include in the plot while keeping the data understandable. It 
would potentially make sense to plot country data individually or by region.


# 3 Joins


Load dataset "flights" and "airports":

```{r}
flights <- nycflights13::flights
head(flights)
flights %>% nrow()

airports <- nycflights13::airports
head(airports)
airports %>% nrow()
```


## 3.1


The following code calculates the average flight delays by destination: 


```{r}
avg_delays_by_dest <- flights %>%
  group_by(dest) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE))

head(avg_delays_by_dest)
avg_delays_by_dest %>% nrow()
```



Join avg_delays_by_dest with airports:

```{r}
avg_delays_airports_join <-
  avg_delays_by_dest %>%
left_join(airports, 
          c("dest" = "faa")) %>%
  filter(dest != "HNL" & dest != "ANC") 

head(avg_delays_airports_join)

ggplot(data = avg_delays_airports_join,
       mapping = aes(lon, lat)) +
    borders("state") +
    geom_point(aes(color = avg_delay)) +
    coord_quickmap() +
  scale_color_gradient2(low = "green", 
                        mid = "light blue", 
                        high = "purple",
                        breaks = c(-20, -10, 0, 10, 20, 30, 40)) +
  labs(title = "Average Flight Delays by Destination",
       subtitle = "United States, 2013",
       x = "longitude",
       y = "latitude",
       color = "Average Delay (min)")
    
```


## 3.2

```{r}

head(flights)

head(airports)

flights_airports_lat_lon <- flights %>%
  left_join(airports,  c("dest" = "faa")) %>%
  left_join(airports, c("origin" = "faa")) %>%
  rename(lat.dest = lat.x,
         lon.dest = lon.x,
         lat.origin = lat.y,
         lon.origin = lon.y)

head(flights_airports_lat_lon)
```



## 3.3


```{r}
flights_airports_lat_lon  %>% 
  group_by(tailnum) %>%
  summarise(n_carriers = n_distinct(carrier)) %>%
  filter(n_carriers > 1) %>%
  nrow()
```

There are 11 planes (identified by tailnumber) that are flown by more than
one carrier. There are seven instances with missing tailnumbers. It is
possible that those missing values would be included in this count.

## 3.4

Question: Is there a relationship between the age of a plane and its delays?


Join flights and planes data sets to create a dataframe with information on
plane metadata and delays:

```{r}
planes <- nycflights13::planes
head(planes)

head(flights)

flights_planes_join <- 
  left_join(flights,
          planes,
          by = "tailnum")

head(flights_planes_join)
```


Plot the relationship between the plane's manufacturing year and arrival and
departure delays:


```{r}
flight_planes_join_avg_delays <-
  flights_planes_join %>%
  group_by(year.y) %>%
  summarise(avg_arr_delay = mean(arr_delay, 
                                 na.rm = TRUE),
            avg_dep_delay = mean(dep_delay, 
                                 na.rm = TRUE),
            n_planes = n()) %>%
  filter(year.y != "NA")

flight_planes_join_avg_delays



ggplot(data = flight_planes_join_avg_delays,
       mapping = aes(y = avg_arr_delay,
                     x = avg_dep_delay)) +
  geom_point(aes(color = year.y,
                 size = n_planes)) +
  scale_color_continuous(type = "viridis") + 
  labs(title = "Average Delays Unaffected by Plane Age",
       x = "Average Departure Delay",
       y = "Average Arrival Delay",
       size = "Number of Planes",
       color = "Year Manufactured")

```

There does not appear to be a relationship between plane age and delays.
There is, as  expected a clear positive relationship between departure and
arrival delays. There is also a greater number of planes that have been 
manufactured more recently than there are older planes in the dataset. 
However, all the planes, regardless of manufacturing year, are grouped in 
the same area of the plot. If anything, the newer planes may have slightly 
higher average delays, but that is likely skewed by there simply being more
newer planes than older planes. 














